Estimate RSV onset and peak timing

### 2023-01-11 RSV_timing_explain/RSV_timing_tutorial.Rmd
update

```{r simulate, echo=T}
set.seed(123)
  n=120 # 10 years
  t <- seq(1,n)
 amp1=2.5 # high amplitude
 freq=1/12 # frequency = 1/period
 amp2=0.5 # low amplitude

@@

#other series shifted by 3 months
 xt2a=amp1*cos(2*3.14159*t*freq+1)

#Simulate some poisson count data
 xt1=rpois(n,exp((1+xt1a)))
 xt2= rpois(n,exp(2+xt2a))
```

@@

```{r, echo=T}
 plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",
      ylab="Cases of RSV", xlab="Month(n)", 
      ylim=range(c(xt1,xt2)))
 lines(t,xt2,col="red")
```

>>>>

```{r simulate, echo=T}
set.seed(123)
n=120 # 10 years
t <- seq(1,n)
amp1=2.5 # high amplitude
freq=1/12 # frequency = 1/period
amp2=2 # low amplitude

@@

#other series shifted by 2 months
xt2a=amp2*cos(2*3.14159*t*freq+1)

#Simulate some poisson count data
xt1=rpois(n,exp((1+xt1a)))
xt2=rpois(n,exp(2+xt2a))
```

@@

```{r, echo=F}
plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",
      ylab="Cases of RSV", xlab="Month(n)", 
      ylim=range(c(xt1,xt2)),
     xlim=c(0,max(t)+10))
lines(t,xt2,col="red")
axis(2, at = seq(0,120,20))
axis(1, at = seq(0,120,12))
text(x=max(t)+8, y=xt1[119], labels = "state1",col="black")
text(x=max(t)+8, y=xt2[119], labels = "state2",col="red")
```

## Investigate the epidemic characteristics of the pseudo-RSV time-serieses

This is based on the prior knowledge that RSV has annual cycle in temperate regions and biannual cycle in high latitude regions. For other viruses or RSV circulation in other climate, you should consider using wavelet analysis to identify the periodicity first.

```{r harm1, echo=TRUE}
# Create the needed harmonic variables with 
# 6 month, 12 month, and 24 month periodicities
t<-1:120
#Create harmonic variables
sin6=sin(2*pi*t/6)
cos6=cos(2*pi*t/6)
sin12=sin(2*pi*t/12)
cos12=cos(2*pi*t/12)
sin24=sin(2*pi*t/24)
cos24=cos(2*pi*t/24)
```

When you write the equations with sin and cos terms, you will need to include them both in an equation. For example, if the coefficient of cos12 is significant, you will need to include sin12 even though the coefficient is not significant in the summary.

## Fit a simple poisson regression with 12 month period for state 1

```{r harm_reg1}
fit1a <- glm(xt1~sin12+cos12, family='poisson')
pred1a<- fitted(fit1a)
#summary(fit1a)
```
```{r ,echo=FALSE}
plot (t, xt1,bty="l", xaxt="n",yaxt="n",col="blue",
      xlab="Month(n)",ylab="Cases of RSV", type="p",
      ylim=range(c(xt1,pred1a)),
     xlim=c(0,max(t)+25),main="State 1")
lines(t,pred1a, lty=1, lwd=3)
axis(2, at = seq(0,120,20))
axis(1, at = seq(0,120,12))
text(x=max(t)+15, y=pred1a[120], labels = "fitted",col="black")
text(x=max(t)+15, y=xt1[120], labels = "observed",col="blue")
```

## Add in 24 month periodicity

\scriptsize
```{r harm_reg2}
fit2a <- glm(xt1~sin12+cos12+sin24+cos24,family='poisson')
pred2a<- fitted(fit2a)
summary(fit2a)
```

## Add in 6 month periodicity

\tiny
```{r harm_reg3}
fit3a<-glm(xt1~sin12+cos12+sin24+cos24+sin6+cos6,family='poisson' )
pred3a<-fitted(fit3a)
summary(fit3a)
```

## Determine best model with AIC (smaller=better)
Winner is Model 1 (12 period). Also, when models have similar AIC score (within 2 points), we prefer a simpler model.
```{r aic}
 AIC(fit1a)
 AIC(fit2a)
 AIC(fit3a)
```

## Fit a simple poisson regression with 12 month period for state 2

```{r harm_reg4}
fit1b <- glm(xt2~sin12+cos12, family='poisson')
pred1b<- fitted(fit1b)
#summary(fit1a)
```

```{r, echo=FALSE}
plot (t,xt2,bty="l", xaxt="n",yaxt="n",col="blue",
      xlab="Month(n)",ylab="Cases of RSV", type="p",
      pch = 20,
      ylim=range(c(xt2,pred1b)),main="State 2")
lines(t,pred1b, lty=1, lwd=1,col="red")
axis(2, at = seq(0,120,20))
axis(1, at = seq(0,120,12))
```

## Calculate amplitudes of the 12 periods in state1 and state2
```{r amp}
beta_sin12_xt1<-coef(fit2a)['sin12']
beta_cos12_xt1<-coef(fit2a)['cos12']

amp12_xt1<-sqrt(beta_sin12_xt1^2+
                  beta_cos12_xt1^2)

amp12_xt1 #True value=2.5

beta_sin12_xt2<-coef(fit1b)['sin12']
beta_cos12_xt2<-coef(fit1b)['cos12']

amp12_xt2<-sqrt(beta_sin12_xt2^2+
                  beta_cos12_xt2^2)

amp12_xt2 #True value=2
```
